---
title: "Introduction to large data management using `R` <br>"
author: "Kostas Mammas, Statistical Programmer <br> mail mammaskon@gmail.com <br>"
date: "EarthBiAs2017, Rhodes Island, Greece"
output:
  github_document:
    toc: true
    toc_depth: 3
always_allow_html: yes
---

<style type="text/css">

body{ /* Normal  */
font-size: 14px;
}
td {  /* Table  */
font-size: 12px;
}
h1 { /* Header 1 */
font-size: 24px;
color: DarkBlue;
}
h2 { /* Header 2 */
font-size: 22px;
color: DarkBlue;
}
h3 { /* Header 3 */
font-size: 18px;
color: DarkBlue;
}
code.r{ /* Code block */
font-size: 12px;
}
pre { /* Code block */
font-size: 12px
}

</style>

# Introduction

The present document works as a user manual of various techniques that can be followed in order to perform efficient data manipulation of environmental data using the `data.table` package.

Before we start you need to setup your working directory by following this [link](https://github.com/mammask/EarthBiAs2017/blob/master/setupR.md).

# Understanding the data structure

The dataset of the precipitation records consists of 5 columns: 

+ **STAID**: Station identifier
+ **SOUID**: Source identifier
+ **DATE** : Date YYYYMMDD
+ **RR**   : Precipitation amount in 0.1 mm
+ **Q_RR** : quality code for RR (0='valid'; 1='suspect'; 9='missing')

The dataset with the available meteorological stations consists of 9 columns:

+ **STAID**: Station identifier
+ **STABANE**: Station name
+ **CN**: Country name
+ **LAT**: Latitude
+ **LONG**: Longitude
+ **HGHT**: Station height
+ **latUpd**: Updated format of station latitude
+ **longUpd**: Updated format of station longitude
+ **LatLong**: Combined format of latitude and longitude


##  Data import

```{r Column-wise and row-wise subsetting, eval = TRUE, warning=FALSE}
# Load libraries
library(data.table)
# Set working directory
setwd("~/Documents/EarthBiAs2017/")
# Read precipitation data
envDat <- readRDS("./data/spanishPrecipRecords.RDS")
# Convert dataset to DT
setDT(envDat)
# Obtain first six rows
head(envDat)
# Obtain column formats
str(envDat)
# Read stations data
stationsData <- readRDS("./data/precipStations.RDS")
# Convert dataset to DT
setDT(stationsData)
```

## Create new variables

In `data.table` it is easy to create new variables or update existing ones using `:=` symbol:

```{r New variables - part 1, eval = TRUE, warning=FALSE}
# Create a new column with the minimum rainfall per station
envDat[Q_RR != 9, maxRainfall := max(RR, na.rm = T), by = STAID]
head(envDat)
```

You can easily delete a variable:

```{r New variables - part 2, eval = TRUE, warning=FALSE}
envDat[, maxRainfall := NULL]
```

## Managing formats

In the previous section we used the `str` function to obtain the dataset formats and it seems that the `DATE` column is a vector of integers. The following script converts `DATE` in date format (`as.Date`):

```{r, eval = TRUE}
envDat[,DATE:=as.Date(as.character(DATE), format("%Y%m%d"))]
```

`:=` is used in order to create a new column or update an existing one.

## Subsetting

In this section, we describe how to select specific rows and columns of a dataset and also make computations based on the subsets:

```{r Subsetting, eval = TRUE, warning=FALSE}
# Select the first 3 columns
envDat[, .(STAID, SOUID, DATE)]
# Select records where precipitation amount is equal to zero
envDat[RR == 0]
# Compute number of rows where precipitation amount is equal to zero
envDat[RR == 0, .N]
# Select specific rows and columns
envDat[RR == 0, .(STAID, SOUID, DATE)]
# Find number of records where Q_RR is equal to 9
envDat[Q_RR == 9, .N]
# Find number of records where Q_RR is equal to 0 or 1 and RR > 0 
envDat[Q_RR != 9 & RR >=0, .N]
# Find number of stations with at least one missing record
envDat[Q_RR == 9, uniqueN(STAID)]
```

## Group by

```{r groub by, eval = TRUE, warning=FALSE}
# Count the number of records by station
envDat[, .N, by = STAID]
# Count the number of records where precipitation amount is greater than 100 mm
# by station
envDat[RR >100, .N, by = STAID]
# Count the number of records where precipitation amount is greater than 100 mm
# by station and available year
envDat[RR >100, .N, by = list(STAID,year(DATE))]
# Find the annual precipitation amount by station where missing records are omitted
envDat[Q_RR !=9, list(N = sum(RR)), by = list(STAID,year(DATE))]
# Perform more than one computations
envDat[Q_RR !=9, list(Total = sum(RR), Mean  = mean(RR)), by = list(STAID,year(DATE))]
```

## Working with `shift`

You can obtain the next or the previous rainfall record by using the `shift` function:

```{r shift, eval = FALSE, warning=FALSE}
# Order rainfall records by station and date
envDat <- envDat[order(STAID,DATE)]
# Obtain DATE lag(1) for each station and fill missing values with -9999
envDat[, shift(x = DATE, fill = NA, n = 1, type = "lag"), by = STAID]
# Obtain RR lag(1) for each station and fill missing values with -9999
envDat[, shift(x = RR, fill = -9999, n = 1, type = "lag"), by = STAID]
# Obtain RR lag(2) for each station and fill missing values with -9999
envDat[, shift(x = RR, fill = -9999, n = 2, type = "lag"), by = STAID]
# Create new columns of lag(1,2) for each station and fill onobserved values
# with -9999
envDat[,`:=` (lag1 = shift(x = RR, fill = -9999, n = 1, type = "lag"),
              lag2 = shift(x = RR, fill = -9999, n = 2, type = "lag")),
       by = STAID]
```

## Creating temporary variables

With `data.table` you can create temporary variables and based on their values you can further compute variables in just one step. In the following example, we create a lag(1) of the daily precipitation record per station and then we compute the mean and minimum precipitation by omitting missing records:

```{r tempVar, eval = TRUE, warning=FALSE}
# Create a temporary variable of lag1 and compute the mean/min of lag1
envDat[Q_RR !=9, { lag1 =  shift(x = RR, fill = NA, n = 1, type = "lag")
                   list( MeanLag1 = mean(lag1, na.rm = T),
                         MinLag1  = min(lag1, na.rm = T))},
       by = STAID]
```

## Special Symbols

### `SD`

`SD` is one of `data.table`'s special symbols and contains a subset of columns. 

```{r special symbols - part 1, eval = TRUE, warning=FALSE}
# Return all columns and all rows
envDat[, .SD]
# Return all columns and first row
envDat[, .SD[1]]
# Return all columns and the first record of each meteorological station
envDat[, .SD[1], by = STAID]
# Return all columns and the first non-missing record of each meteorological station
envDat[Q_RR != 9, .SD[1], by = STAID]
# Return specific columns (station id, date and rainfall record)
envDat[,.SD, .SDcols = c("STAID","DATE", "RR")]
```

It is very convenient to perform dynamic computations across many variables:

```{r special symbols - part 2, eval = TRUE, warning=FALSE}
# Perform dynamic computations accross variables
envDat[, lapply(.SD, min), .SDcols = c("RR")]
# Perform dynamic computations accross variables and by station
envDat[, lapply(.SD, min), .SDcols = c("RR"), by = "STAID"]
# Perform dynamic computations accross variables and by station
# where rainfall records are non missing
envDat[Q_RR !=9, lapply(.SD, min), .SDcols = c("RR"), by = "STAID"]
# More variables can be added in SD
envDat[Q_RR !=9, lapply(.SD, min), .SDcols = c("RR", "Q_RR"), by = "STAID"]
```

### `.N`

`.N` is a special in-built variable that holds the number of observations in the current group. When we group by`origin`, `.N` returns the number of rows of the dataset:

```{r special symbols - part 3, eval = FALSE, warning=FALSE}
# Return the number of records per station
envDat[, .N, by = "STAID"]
# Create a counter variable per station
envDat[, 1:.N, by = "STAID"]
```

`SD` can be combined with `.N` to access specific rows of the dataset:

```{r special symbols - part 4, eval = FALSE, warning=FALSE}
# Return last record of each one of the three variables
envDat[,.SD[.N], .SDcols = c("STAID","DATE", "RR")]
# Return all but the last record for the selected columns
envDat[,.SD[1:(.N-1)], .SDcols = c("STAID","DATE", "RR")]
# Return all but the last record for the selected columns by station
envDat[,.SD[1:(.N-1)], .SDcols = c("STAID","DATE", "RR"), by = "STAID"]
```

### Exercise 1: Summary statistics

As a first task you will have to pick randomly 10 meteorological stations and compute the following metrics:

+ Mean, Min, Max, Median and Total rainfall amount (missing records are omitted)
+ Compute the number of missing records per station
+ Which is the month with the highest number of missing values per station?

## `Join` using `data.table`

Before we provide examples about how to merge using the `data.table` approach please go through the following examples:

+ Joining data in R using `data.table` ([link](https://rstudio-pubs-static.s3.amazonaws.com/52230_5ae0d25125b544caab32f75f0360e775.html)).

In the following example we will compute the number of records per station and obtain the station name:

```{r Join - part 1, eval = TRUE, warning=FALSE}
# Count the number of records per station
recsPerStat <- envDat[, .N, by = STAID]
# Define key for both datasets
setkey(recsPerStat, STAID)
setkey(stationsData, STAID)
# Obtain the station name from the stations table
recsPerStat[stationsData[, list(STAID, STANAME)], nomatch = 0L]
# Obtain station with the most rainfall records - inner join
recsPerStat[stationsData[, list(STAID, STANAME)], nomatch = 0L][order(N, decreasing = T)]
```

In the next lecture we will find metrics for various stations and based on the stations' location we will create maps.

## Environmental indices with `data.table`

### Number of days with "extreme" rainfall events

Extreme rainfall event is defined as a daily rainfall record exceeding a specific rainfall amount. In this case, we calculate the probability distribution of the daily rainfall series and we define extreme records those which exceed the 90% of the values.

```{r extreme rainfall events - part 1, eval = TRUE, warning=FALSE}
# Calculate extreme rainfall events
# - Calculate the 90% of the values of the probability 
#   distribution of daily rainfall
# - Find the records who exceed this threshold
# - Count the records who exceed this threshold
# - Make this computation for each station and each available year
envDat[Q_RR != 9, {extrVal     = quantile(RR, probs = 0.9)[[1]]
                   ExtrEv      = RR[RR > extrVal]
                   list(N = length(ExtrEv))
                   },
       by = list(STAID, year(DATE))]
```

In the previous example we included also the days with no rainfall amount. We can calculate the 90% of the probability distribution for the days with rainfall amount greater than zero:

```{r extreme rainfall events - part 2, eval = TRUE, warning=FALSE}
# Calculate extreme rainfall events
# - Calculate the 90% of the values of the probability 
#   distribution of daily rainfall
# - Find the records who exceed this threshold
# - Count the records who exceed this threshold
# - Make this computation for each station and each available year
envDat[Q_RR != 9, {extrVal     = quantile(RR[RR>0], probs = 0.9)[[1]]
                   ExtrEv      = RR[RR > extrVal]
                   list(N = length(ExtrEv))
                   },
       by = list(STAID, year(DATE))]
```

### Maximum number of consecutive rainfall events

In the following example, we compute the maximum number of consecutive rainfall days for each station and available year. 

```{r consecutive rainfall events - part 1, eval = TRUE, warning=FALSE}
# Compute maximum number of consecutive daily rainfall events per station
# Temporary variables: idx        - indexing days with rainfall
#                      diff       - find if the current day and the next day are days with rainfall
#                      lagdiff    - shift series on position
#                      startPoint - create a starting point
envDat[, {idx        = 1*(RR>0);
          diff       = shift(x = idx, fill = 0, type = "lead", n=1) - idx;
          lagdiff    = c(0,diff[-.N])
          startPoint = 1*(lagdiff<0)
          list(DATE  = DATE,
               RR    = RR,
               group =cumsum(startPoint))},
          by = list(STAID, year(DATE))
      ][RR != 0, .N, by = list(STAID, year, group)][, list(Max = max(N)),
                                                                by = list(STAID, year)]
```

This example can be easily extended so as to compute the number of consecutive days per station, year and month.

```{r consecutive rainfall events - part 2, eval = TRUE, warning=FALSE}
# Compute maximum number of consecutive daily rainfall events per station
# Temporary variables: idx        - indexing days with rainfall
#                      diff       - find if the current day and the next day are days with rainfall
#                      lagdiff    - shift series one position
#                      startPoint - create a starting point
envDat[, {idx        = 1*(RR>0);
          diff       = shift(x = idx, fill = 0, type = "lead", n=1) - idx;
          lagdiff    = c(0,diff[-.N])
          startPoint = 1*(lagdiff<0)
          list(DATE  = DATE,
               RR    = RR,
               group =cumsum(startPoint))},
          by = list(STAID, year(DATE), month(DATE))
      ][RR > 0, .N, by = list(STAID, year, month, group)][, list(Max = max(N)),
                                                                by = list(STAID, year, month)]
```

Also, this example can be easily extended in order to compute the maximum number of consecutive drought days.

### Exercise 2: Compute the maximum number of extreme consecutive rainfall events

You will need to count the maximum number of consecutive extreme rainfall events by station and by year considering that an extreme rainfall event is a record with value greater than the 90% of the distribution of daily rainfall of each station and year. 

# Visualization of environmental rainfall series

## Visualization of rainfall series - Annual rainfall amount

In the following example we compute the annual rainfall amount for a specific meteorological station:

```{r Visualizing rainfall series - annual rainfall - part 1, eval = TRUE, warning=FALSE, fig.width=9, fig.height=3}
# Load dygraphs
library(dygraphs)
statID    <- envDat[,unique(STAID)][1]
annualSer <- envDat[STAID == statID & Q_RR != 9, list(N = sum(RR)), by = year(DATE)]
# Produce interactive plot of annual rainfall series
p <- dygraph(annualSer, main = paste0("Annual rainfall series of station: ",statID)) %>%
     dyAxis("y", label = "Rainfall amount (mm)") %>% dyAxis("x", label = "Year")
# p 
```

We can add more features in the previous plot:

```{r Visualizing rainfall series - annual rainfall - part 2, eval = TRUE, warning=FALSE, fig.width=9, fig.height=3}
# Produce interactive plot of annual rainfall series
p <- dygraph(annualSer, main = paste0("Annual rainfall series of station: ",statID)) %>%
     dyAxis("y", label = "Rainfall amount (mm)") %>% dyAxis("x", label = "Year") %>% 
     dyRangeSelector() %>% dyOptions(fillGraph = TRUE, fillAlpha = 0.4)
# p
```

In the following example we will compute a chart for each meteorological station using `data.table`:

```{r Visualizing rainfall series - annual rainfall - part 3, eval = TRUE, warning=FALSE, fig.width=9, fig.height=3}

# Function to produce plot of annual rainfall based on the station ID, year and rainfall amount
annualPlot <- function(year, N, ID){
  
  # function name: annualPlot
  #       purpose: to create annual rainfall series plots for each meteorological station
  #        inputs: year   - a vector with the available years
  #                N      - a vector with the annual rainfall series
  #                ID     - the station ID
metDat <- data.table(year, N)
p <- dygraph(metDat, main = paste0("Annual rainfall series of station: ",ID)) %>%
     dyAxis("y", label = "Rainfall amount (mm)") %>% dyAxis("x", label = "Year") %>% 
     dyRangeSelector() %>% dyOptions(fillGraph = TRUE, fillAlpha = 0.4)
  
  return(p)
}

# Produce annual rainfall series
anRainfall <- envDat[Q_RR != 9, list(N = sum(RR)), by = list(STAID, year(DATE))]
# Produce all the available plots
plotList <- anRainfall[, list(Plot = list(annualPlot(year, N, STAID))), by = STAID]
# plotList
```

How can we access a specific plot?

### Exercise 3: - Produce plots for other summary statistics

+ Produce plots of the maximum number of consecutive rainfall events for each meteorological station and available year.

+ Can you replace the station ID by the station name at the title of the plots?

## Calculation of annual rainfall trends for a specific station

```{r calculation of rainfall trend - one station, eval = TRUE, warning=FALSE, fig.width=9, fig.height=3}
trendDat <- annualSer[, { trendModel = lm(data = data.table(year, N), formula = "N~1+year")
                          trendline  = coef(trendModel)[[2]] * year + coef(trendModel)[[1]]
                          list(year = year, N = N, trend = trendline)
                         }
          ]
head(trendDat)
```

### Plot of annual rainfall trends for a specific station

```{r plot rainfall trend - one station, eval = TRUE, warning=FALSE, fig.width=9, fig.height=3}

p <- dygraph(trendDat, main = paste0("Annual rainfall series - Station: ", statID)) %>%
     dyAxis("y", label = "N") %>% dyAxis("x", label = "year") %>% 
     dySeries("trend", stepPlot = FALSE, fillGraph = FALSE) %>%
     dyRangeSelector() %>% dyOptions(fillGraph = FALSE, fillAlpha = 0.4)
# p
```


## Calculation of annual rainfall trends for all the available rainfall stations

In the following example we compute the annual rainfall amount for each meteorological station and then annual rainfall trend per station is calculated:

```{r calculation of rainfall trend - all stations, eval = TRUE, warning=FALSE, fig.width=9, fig.height=3}
# Calculate annual rainfall per station
annualSer <- envDat[Q_RR != 9, list(N = sum(RR)), by = list(STAID, year(DATE))]
# Calculate annual rainfall trend for each station
trendDat  <- annualSer[, { trendModel = lm(data = data.table(year, N), formula = "N~1+year")
                           trendline  = coef(trendModel)[[2]] * year + coef(trendModel)[[1]]
                           list(year = year, N = N, trend = trendline)
                         },
                       by = STAID
          ]
head(trendDat)
```

<!-- ## Spatial visualization -->

<!-- In this section we provide visualization outputs using the `leaflet` package. The following script provides the location of the meteorological station of Rhodes: -->


<!-- ```{r Spatial visualization - part 1, eval = TRUE, warning=FALSE, fig.width=9.5, fig.height=4} -->
<!-- # Load leaflet library -->
<!-- library(leaflet) -->
<!-- # Obtain information about Rhodes meteorological station -->
<!-- statInfo <- stationsData[STANAME == "RHODOS"] -->
<!-- # Create leaflet map using the station coordinates -->
<!-- m <- leaflet() %>% setView(lng =statInfo[,longUpd],  lat = statInfo[,latUpd], zoom = 7) -->
<!-- m %>% addTiles()  %>% addMarkers(lng =statInfo[,longUpd],  lat = statInfo[,latUpd],  -->
<!--                                  popup="Rhodes Meteorological Station") -->
<!-- ``` -->

<!-- <br> -->
<!-- Following the same logic we can produce maps of all the available meteorological stations in Greece: -->

<!-- ```{r  Spatial visualization - part 2, eval = TRUE, warning=FALSE, fig.width=9.5, fig.height=4} -->
<!-- greekStations <- stationsData[CN=="GR"] -->
<!-- # Create leaflet map using the stations coordinates -->
<!-- m <- leaflet() %>% setView(lng =greekStations[,mean(longUpd)],  lat = greekStations[,mean(latUpd)], zoom = 6) -->
<!-- m %>% addTiles()  %>% addMarkers(lng = greekStations[,longUpd],  lat = greekStations[,latUpd],  -->
<!--                                  popup = greekStations[,STANAME]) -->
<!-- ``` -->

<!-- <br> -->
<!-- In the following example we will incorporate information about the height of each meteorological station at each popup: -->

<!-- ```{r  Spatial visualization - part 3, eval = TRUE, fig.width=9.5, fig.height=4, message=FALSE, warning=FALSE, results='hide'} -->
<!-- # Load library stringr -->
<!-- library(stringr) -->
<!-- greekStations <- stationsData[CN=="GR"] -->
<!-- greekStations[, Description:= paste0("Station: ",str_to_title(STANAME), "<br> Height: ", HGHT, " m")] -->
<!-- ``` -->

<!-- ```{r  Spatial visualization - part 4, eval = TRUE, warning=FALSE, fig.width=9.5, fig.height=4} -->
<!-- # Create leaflet map using the stations coordinates -->
<!-- m <- leaflet() %>% setView(lng =greekStations[,mean(longUpd)],  lat = greekStations[,mean(latUpd)], zoom = 6) -->
<!-- m %>% addTiles()  %>% addMarkers(lng = greekStations[,longUpd],  lat = greekStations[,latUpd],  -->
<!--                                  popup = greekStations[,Description]) -->
<!-- ``` -->

<!-- ### Exercise 4: Develop a spatial map -->

<!-- + Create a map of all the Spanish meteorological stations including the height of each station. -->